PanGraph: scalable bacterial pan-genome graph construction

### 2022-09-28 main.tex
feat: improved projection figures
feat: modified main

@@

\subsection{Graph Export and Availability}
The constructed pangenome graph can be exported in a variety of file formats for downstream analysis and visualization.
In addition to a custom JSON schema, PanGraph can export the alignment as a GFA file, where each \emph{pancontig} is represented as a segment and each genome as a path.
This allows for visualization in software such as Bandage \cite{wick2015bandage}.
Lastly, we provide functionality to export as a conventional pangenome, however wherein \emph{pancontigs} supplant putative gene clusters, that can be visualized directly by the PanX toolkit \cite{ding2018panx}.

PanGraph is published under an MIT license with source code, extensive documentation, examples, and instructions for installation available at \url{github.com/neherlab/pangraph}.
Additionally, pre-built binaries are available as GitHub releases.
All data and scripts used to validate PanGraph are available within the same repository.

@@

In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $10^{-2}$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line toolsuite for external use.

@@

We quantified the accuracy of PanGraph algorithm by computing the distance, $\delta$, of the inferred breakpoints relative to their known locus stored from the evolutionary simulation.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $r$ and thus underlying graph complexity.
The predominant determinant of accuracy was the sample diversity, shown in Fig. \ref{fig:toy-accuracy} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
    \caption{{\bf Algorithm performance}.
        PanGraph scales linearly with the number of input genomes.
        This is a direct result of the guide tree simplification.
        The solid line and ribbons display the mean and standard deviation respectively.
        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
    }
    \label{fig:toy-performance}
\end{figure}

The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.

@@

\begin{figure*}[htb]
    \includegraphics[width=.8\textwidth]{figs/proj_single_escherichia_coli.pdf}
    \caption{{\bf Test of graph marginalization}.}
    \label{fig:marginalization}
\end{figure*}

>>>>

% TODO: check consistency of numberings of figures and sections between SI and main text

@@

\subsection{Graph Export and Availability}
The constructed pangenome graph can be exported in a variety of file formats for downstream analysis and visualization.
In addition to a custom JSON schema, PanGraph can export the alignment as a GFA file, where each \emph{pancontig} is represented as a segment and each genome as a path.
This allows for visualization in software such as Bandage \cite{wick2015bandage}.
Lastly, we provide functionality to export as a conventional pangenome, however wherein \emph{pancontigs} supplant putative gene clusters, that can be visualized directly by the PanX toolkit \cite{ding2018panx}.
% TODO: add GFA version?

PanGraph is published under an MIT license with source code, extensive documentation, examples, and instructions for installation available at \url{github.com/neherlab/pangraph}.
Additionally, pre-built binaries are available as GitHub releases.
All data and scripts used to validate PanGraph are available within the same repository.
% TODO: add data?

@@

As a first validation step we use generated synthetic data to quantify the performance characteristics of \textit{PanGraph} as a function of input size, and its accuracy as a function of sequence diversity.

\subsubsection*{Generation of synthetic data}

We simulated populations of size $N=100$ and sequence length $L=50\,000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI Section I for details).
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line toolsuite for external use.

\subsubsection*{Performances as a function of dataset size}

The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length, and generating pangenome graphs for the resulting sequences.
The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
Benchamrks of CPU and memory usage are reported in SI Section IIA.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
    \caption{{\bf Algorithm performance}.
        PanGraph scales linearly with the number of input genomes.
        This is a direct result of the guide tree simplification.
        The solid line and ribbons display the mean and standard deviation over 50 runs.
        All runs were performed utilizing 8 cores.
    }
    \label{fig:toy-performance}
\end{figure}

\subsubsection*{Accuracy as a function of sequence divergence}

We quantified the accuracy of PanGraph algorithm by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf. SI Section IIB for details). We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and HGT $h$. For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel, namely \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} option and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $h$ and thus underlying graph complexity. The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf. SI Fig.~2). For low-diversity samples breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf. SI Fig.~3). The choice of alignment kernel influences the threshold diversity at which accuracy is lost.

%  TODO: talk aboit the figure
For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.

@@

% \subsection{Validation on synthetic data}

% In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
% Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
% In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $10^{-2}$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
% The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
% The simulation framework is distributed within the PanGraph command line toolsuite for external use.

% \begin{figure}[htb]
%     \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
%     \caption{{\bf Accuracy against synthetic data}.
%         We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} and \textit{asm20} option, and \textit{mmseqs2}. In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp. The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of 100bp, as a function of average pairwise sequence divergence. Line and shaded area represents mean and standard deviation over 25 repetitions.
%         Graph reconstruction with \textit{minimap2} and \textit{asm20} is accurate up to around 5\% average sequence divergence, while for \textit{mmseqs2} accuracy is maintained until $\sim$12\% divergence.
%     }
%     \label{fig:toy-accuracy}
% \end{figure}

% We quantified the accuracy of PanGraph algorithm by computing the distance, $\delta$, of the inferred breakpoints relative to their known locus stored from the evolutionary simulation.
% Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $r$ and thus underlying graph complexity.
% The predominant determinant of accuracy was the sample diversity, shown in Fig. \ref{fig:toy-accuracy} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
% For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
% Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
% This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.

% \begin{figure}[htb]
%     \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
%     \caption{{\bf Algorithm performance}.
%         PanGraph scales linearly with the number of input genomes.
%         This is a direct result of the guide tree simplification.
%         The solid line and ribbons display the mean and standard deviation respectively.
%         All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
%     }
%     \label{fig:toy-performance}
% \end{figure}

% The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
% The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
% Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
% We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.

@@

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
    \caption{{\bf Test of graph marginalization}.}
    \label{fig:marginalization}
\end{figure*}
